A Large-Scale Study of the Use of Eval in R

### 2020-08-25 inst/analysis/corpus-analysis.Rmd
V.A

```{r setup, include=FALSE}
for (x in c(
  "dplyr", 
  "DT",
  "fs",
  "fst",
  "ggplot2",
  "ggthemes",
  "lubridate",
  "readr",
  "purrr",
  "stringr",
  "tidyr",
  "viridis"
  )) {
  suppressPackageStartupMessages(library(x, character.only=TRUE))
}

@@

```{r load corpus}
corpus <- 
  read_fst(CORPUS_DETAILS_FILE) %>%
  as_tibble() %>%
  mutate_at(
    vars(
      starts_with("runnable_code"),
      starts_with("runnable_files"),
      starts_with("package_"),
      starts_with("funs_"),
    ), 
    replace_na, 
    0
  )
```

@@

```{r summary}
corpus_runtime  <- corpus$runnable_time / corpus$runnable_files
runnable_code   <- corpus$runnable_code
corpus_programs_files <- corpus$runnable_files

overview_table(
  r("packages",             corpus),
  r("in cranlogs",          filter(corpus, !is.na(rank))),
  r("package programs",     sum(corpus_programs_files)),

  r("runnable code",        sum(runnable_code)),
  r("runnable code mean",   mean(runnable_code)),
  r("runnable code median", median(runnable_code)),

  r("examples code",        sum(corpus$runnable_code_examples)),
  r("tests code",           sum(corpus$runnable_code_tests)),
  r("vignettes code",       sum(corpus$runnable_code_vignettes)),

@@

  r("code",                 sum(corpus$package_code)),
  r("R code",               sum(corpus$package_r_code)),
  r("native code",          sum(corpus$package_native_code)),

  r("functions",            sum(corpus$funs_public+corpus$funs_private)),
  r("function with eval",   sum(corpus$funs_with_eval)),
  r("public functions",     sum(corpus$funs_public)),
  r("private functions",    sum(corpus$funs_private)),

  r("mean revdes",          mean(corpus$revdeps)),
  r("median revdes",        median(corpus$revdeps)),

  r("runtime",              as.duration(sum(corpus$runnable_time))),
  r("mean runtime",         as.duration(mean(corpus_runtime))),
  r("median runtime",       as.duration(median(corpus_runtime)))
)
```

```{r generals}
overview_table(
  r("all cran", dim(available.packages())[1]),
  r("core packages", length(CORE_PACKAGES))
)
```

@@

```{r plot corpus}
coverage_mean   <- mean(corpus$coverage_expr)
coverage_median <- median(corpus$coverage_expr)
revdeps_mean    <- mean(corpus$revdeps)
revdeps_median  <- median(corpus$revdeps)

corpus %>%
  mutate(
    outlier=package_code > PACKAGE_SIZE_OUTLIER,
    label=ifelse(rank <= 25 | outlier, package, NA)  #if_else(outlier, str_glue("{package}"), as.character(NA))
  ) %>%
  ggplot(
    aes(
      x=coverage_expr, 
      y=revdeps,
      label=label,
      size=package_code
    )
  ) +
  geom_point(alpha=.6) +
  geom_text(size=2, check_overlap = T, vjust=1.5, na.rm = TRUE) +

  geom_hline(aes(yintercept=revdeps_mean), linetype="dashed", color="red", size=0.2) +
  geom_vline(aes(xintercept=coverage_mean), linetype="dashed", color="red", size=0.2) +

  scale_x_continuous(labels = scales::percent) + 
  scale_y_log10(labels = scales::comma) +
  scale_size(range = c(.1, 3), labels = scales::comma) +
  theme(
    legend.position=c(0.2, 0.85),
    legend.box="horizontal",
    legend.box.background = element_rect(fill="white", size=0.1)
  ) +
  labs(
    x="Code coverage", 
    y="Number of reverse dependencies (log)", 
    size="Code size (R + native)",
    linetype=""
  ) +
  guides(
    size=guide_legend(order=1),
    color=guide_legend(order=2),
    linetype=F
  )

@@

```{r}
kaggle_runnable <- filter(kaggle, runnable)
kaggle_finished <- filter(kaggle, exitval==0)
kaggle_code     <- kaggle_finished$code
kaggle_runtime  <- kaggle_finished$runtime

overview_table(
  r("kaggle",                 kaggle),
  r("runnable kaggle",        kaggle_runnable),
  r("finished kaggle",        kaggle_finished),
  r("finished kaggle code",   sum(kaggle_code)),
  r("finished kaggle mean",   mean(kaggle_code)),
  r("finished kaggle median", median(kaggle_code)),
  r("kaggle runtime",         as.duration(sum(kaggle_runtime))),
  r("kaggle mean",            as.duration(mean(kaggle_runtime))),
  r("kaggle median",          as.duration(median(kaggle_runtime)))
)
```

@@

```{r}
overview_table(
  r("all programs", nrow(kaggle_finished) + sum(corpus_programs_files)),
  r("all runnbale code", sum(kaggle_code) + sum(runnable_code))
)
```

```{r}
corpus_programs <-
  left_join(
    filter(runs, exitval==0) %>% select(package, file, time), 
    runnable_code_metadata %>% select(package, file, code),
    by=c("package", "file")
  ) %>%
  transmute(id=paste(package, ":", file), code, runtime=time, corpus="packages")

kaggle_programs <-
  kaggle %>%
  filter(exitval==0) %>%
  transmute(id, code, runtime=as.numeric(runtime, units="seconds"), corpus="kaggle")

programs <- bind_rows(corpus_programs, kaggle_programs)
```

```{r}
programs %>%
  ggplot(aes(y=code, x=corpus, fill=corpus)) + 
  geom_flat_violin(trim=FALSE, width=.5) +
  geom_dotplot(
    binaxis="y", 
    dotsize=0.1, 
    stackdir="down", 
    binwidth=0.01, 
    position=position_nudge(-0.025)) +
  scale_y_log10(labels = scales::comma) +
  scale_x_discrete(labels=c("kaggle"="Kaggle", "packages"="Packages")) +
  scale_fill_viridis(discrete = TRUE) +
  guides(fill = FALSE) +
  labs(y="Code size (log)") +
  theme(
    legend.box="horizontal",
    axis.title.x=element_blank()
  )

@@

```{r}
programs %>%
  ggplot(aes(y=runtime, x=corpus, fill=corpus)) + 
  geom_flat_violin(trim=FALSE, width=.5) +
  geom_dotplot(
    binaxis="y", 
    dotsize=0.1, 
    stackdir="down", 
    binwidth=0.01, 
    position=position_nudge(-0.025)) +
  scale_y_log10(labels = function(x, ...) {
    min <- as.integer(x/60L)
    sec <- as.integer(x%%60L)
    sprintf("%.2d:%.2d", min, sec)
  }) +
  scale_x_discrete(labels=c("kaggle"="Kaggle", "packages"="Packages")) +
  scale_fill_viridis(discrete = TRUE) +
  guides(fill = FALSE) +
  labs(y="Running time [mm:ss] (log)") +
  theme(
    legend.box="horizontal",
    axis.title.x=element_blank()
  )
ggsave(path(PLOT_DIR, "programs-runtime.pdf"), width=3.5, height=5.0)
```

>>>>

```{r setup, include=FALSE}
local(for (x in c(
  "dplyr", 
  "DT",
  "fs",
  "fst",
  "ggplot2",
  "ggthemes",
  "lubridate",
  "readr",
  "purrr",
  "stringr",
  "tidyr",
  "viridis"
  )) {
  suppressPackageStartupMessages(library(x, character.only=TRUE))
})

@@

```{r load corpus}
corpus <- 
  read_fst(CORPUS_DETAILS_FILE) %>%
  as_tibble() %>%
  mutate_at(
    vars(
      starts_with("runnable_code"),
      starts_with("runnable_files"),
      starts_with("package_"),
      starts_with("funs_"),
      "evals"
    ), 
    replace_na, 
    0
  )
```

@@

```{r compute corpus programs}
corpus_programs <-
  left_join(
    filter(runs, exitval==0) %>% select(package, file, time), 
    runnable_code_metadata %>% select(package, file, code),
    by=c("package", "file")
  ) %>%
  transmute(
    package,
    file,
    type=basename(dirname(file)),
    code, 
    runtime=time
  )
```

There will be some programs that were not strictly part of examples, tests or vignettes, or they do not have
counts for code size or for execution time.

```{r compute program inconsistencies}
corpus_programs_inconsistent <-
  corpus_programs %>%
  filter(
    !(type %in% c("examples", "tests", "vignettes")) |
    is.na(code) | 
    is.na(runtime) |
    code == 0
  )

corpus_programs_inconsistent %>% 
    my_datatable()
```

Since, there are just a few of them, we ignore them.

```{r filter out inconsistent programs}
corpus_programs <- anti_join(corpus_programs, corpus_programs_inconsistent, by=c("package", "file"))
```

```{r summary}
runnable_code <- 
  corpus_programs %>%
  mutate(files=1) %>%
  pivot_wider(names_from=type, values_from=c(files, code)) %>%
  select(-file) %>%
  group_by(package) %>%
  summarise_all(sum, na.rm=TRUE) %>%
  group_by(package) %>%
  mutate(
    files=files_examples + files_tests + files_vignettes,
    code=code_examples + code_tests + code_vignettes
  )

rs <- function(name, value) {
  stopifnot(is.numeric(value))
  bind_rows(
    r(name, sum(value)),
    r(paste(name, "mean"), mean(value)),
    r(paste(name, "median"), median(value)),
    r(paste(name, "min"), min(value)),
    r(paste(name, "max"), max(value))
  )
}

corpus_program_examples <- filter(corpus_programs, type=="examples")
corpus_programs_tests <- filter(corpus_programs, type=="tests")
corpus_programs_vignettes <- filter(corpus_programs, type=="vignettes")

overview_table(
  r("packages",    corpus),
  rs("revdeps", corpus$revdeps),
  r("in cranlogs", filter(corpus, !is.na(rank))),

  rs("package programs",         runnable_code$files),
  rs("package programs code",    runnable_code$code),
  rs("package programs runtime", as.duration(runnable_code$runtime)),

  rs("examples programs",  runnable_code$files_examples),
  rs("tests programs",     runnable_code$files_tests),
  rs("vignettes programs", runnable_code$files_vignettes),

  rs("examples code",  corpus_program_examples$code),
  rs("tests code",     corpus_programs_tests$code),
  rs("vignettes code", corpus_programs_vignettes$code),

  rs("examples runtime",  corpus_program_examples$runtime),
  rs("tests runtime",     corpus_programs_tests$runtime),
  rs("vignettes runtime", corpus_programs_vignettes$runtime),

@@

  rs("code",        corpus$package_code),
  rs("R code",      corpus$package_r_code),
  rs("native code", corpus$package_native_code),

  rs("functions",          corpus$funs_public + corpus$funs_private),
  rs("function with eval", corpus$funs_with_eval),
  rs("public functions",   corpus$funs_public),
  rs("private functions",  corpus$funs_private),

  rs("revdes", corpus$revdeps)
)
```

```{r generals}
overview_table(
  r("all cran", dim(available.packages())[1]),
  r("core packages", length(CORE_PACKAGES)),
  r("most downloaded treshold", MOST_DOWNLOADED),
  r("large package treshold", PACKAGE_SIZE_OUTLIER)
)
```

@@

```{r plot corpus}
coverage_mean   <- mean(corpus$coverage_expr)
coverage_median <- median(corpus$coverage_expr)
revdeps_mean    <- mean(corpus$revdeps)
revdeps_median  <- median(corpus$revdeps)
funs_with_eval_mean    <- mean(corpus$funs_with_eval)
funs_with_eval_median  <- median(corpus$funs_with_eval)

size_bins       <- c(0L, 1e3L, 5e3L, 1e4L, 5e4L, 1e5L)
size_bins_lasb  <- c("< 1000", "1,000 - 4,999", "5,000 - 9,999", "10,000 - 49,999", "> 50,000")

corpus %>%
  mutate(
    outlier=funs_with_eval > 30 | package_r_code > PACKAGE_SIZE_OUTLIER,
    label=ifelse(rank <= MOST_DOWNLOADED | outlier, package, NA),
    code_size=cut(package_r_code, size_bins, labels=size_bins_lasb, include.lowest=TRUE)
  ) %>%
  ggplot(
    aes(
      x=coverage_expr, 
#      y=funs_with_eval,
      y=evals,
      label=label,
      size=code_size
    )
  ) +
  geom_point(alpha=.4) +
  geom_text(size=3, check_overlap = T, vjust=-1.5, na.rm = TRUE) + 
  geom_hline(aes(yintercept=funs_with_eval_mean), linetype="dashed", color="red", size=0.2) +
  geom_vline(aes(xintercept=coverage_mean), linetype="dashed", color="red", size=0.2) +
  scale_x_continuous(labels = scales::percent) + 
  scale_y_continuous(trans=scales::pseudo_log_trans(), breaks=c(0, 1, 2, 5, 10, 20, 50, 100, 200, 500)) +
  scale_size_discrete(range = c(1, 6)) +
  theme(
    legend.position=c(0.2, 0.85),
    legend.box="horizontal",
    legend.box.background = element_rect(fill="white", size=0.1)
  ) +
  labs(
    x="Code coverage", 
    y="Number of eval call sites", 
    size="Lines of R code",
    linetype=""
  ) +
  guides(
    size=guide_legend(order=1),
    color=FALSE,
    linetype=FALSE
  )

@@

```{r compute kaggle programs}
kaggle_programs <-
  kaggle %>%
  filter(exitval==0) %>%
  transmute(
    id, 
    code, 
    runtime=as.numeric(runtime, units="seconds")
  )
```

```{r kaggle summary}
kaggle_code     <- kaggle_programs$code
kaggle_runtime  <- kaggle_programs$runtime

overview_table(
  r("kaggle",                 kaggle),
  r("runnable kaggle",        filter(kaggle, runnable)),
  r("finished kaggle",        kaggle_programs),
  r("finished kaggle code",   sum(kaggle_code)),
  r("finished kaggle mean",   mean(kaggle_code)),
  r("finished kaggle median", median(kaggle_code)),
  r("kaggle runtime",         as.duration(sum(kaggle_runtime))),
  r("kaggle mean",            as.duration(mean(kaggle_runtime))),
  r("kaggle median",          as.duration(median(kaggle_runtime)))
)
```

@@

```{r compute programs}
programs <- bind_rows(
  corpus_programs %>% mutate(corpus="packages") %>% select(code, runtime, corpus),
  kaggle_programs %>% mutate(corpus="kaggle") %>% select(code, runtime, corpus)
)
```

```{r all summary}
overview_table(
  r("all programs",      nrow(kaggle_programs) + nrow(corpus_programs)),
  r("all runnbale code", sum(kaggle_code) + sum(runnable_code$code)),
  r("all code",          sum(kaggle_code) + sum(corpus$package_code) + sum(runnable_code$code))
)
```

```{r programs-code, eval=F}
programs %>%
  ggplot(aes(y=code, x=corpus, fill=corpus)) + 
  geom_violin() +
  scale_y_log10(labels = scales::comma) +
  scale_x_discrete(labels=c("kaggle"="Kaggle", "packages"="Packages")) +
  scale_fill_viridis(discrete = TRUE) +
  guides(fill = FALSE) +
  labs(y="Code size (log)") +
  theme(
    legend.box="horizontal",
    axis.title.x=element_blank()
  )

@@

```{r programs-runtime, eval=F}
programs %>%
  ggplot(aes(y=runtime, x=corpus, fill=corpus)) + 
  geom_flat_violin(trim=FALSE, width=.5) +
  geom_dotplot(
    binaxis="y", 
    dotsize=0.1, 
    stackdir="down", 
    binwidth=0.01, 
    position=position_nudge(-0.025)) +
  scale_y_log10(labels = function(x, ...) {
    min <- as.integer(x/60L)
    sec <- as.integer(x%%60L)
    sprintf("%.2d:%.2d", min, sec)
  }) +
  scale_x_discrete(labels=c("kaggle"="Kaggle", "packages"="Packages")) +
  scale_fill_viridis(discrete = TRUE) +
  guides(fill = FALSE) +
  labs(y="Running time [mm:ss] (log)") +
  theme(
    legend.box="horizontal",
    axis.title.x=element_blank()
  )
ggsave(path(PLOT_DIR, "programs-runtime.pdf"), width=3.5, height=5.0)
```

### 2020-08-25 inst/paper/main.tex
V.A

\documentclass[conference]{IEEEtran}
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{listings}
\newcommand{\missingNumber}{\textcolor{red}{XX}\xspace}
\newcommand{\missingPercentage}{\textcolor{red}{XX\%}\xspace}
\newcommand{\missingTable}{\textcolor{red}{XXTable}\xspace}
\newcommand{\missingGraph}{\textcolor{red}{XXGraph}\xspace}
\usepackage[usenames,dvipsnames]{xcolor} 
\usepackage{xspace}
\usepackage{url}

@@

\newcommand{\eg}{\emph{e.g.},\xspace}
\newcommand{\ie}{\emph{i.e.},\xspace}
\newcommand{\cf}{\emph{cf.}\xspace}

@@

\begin{figure*}[!tb]\centering\includegraphics[width=.8\linewidth]
  {corpus.pdf}\caption{Corpus}\label{fig:corpus}
\end{figure*}

@@

We have selected the TOP \CorpusPackages packages based on the number of reverse
dependencies, \ie how many packages depend on them\footnote{extracted from
  package metadata using builtin function}. Using the number of reverse
dependencies helps us to choose popular packages which in general should be of a
better quality and should have better code coverage. The resulting set consists
of \CorpusRCodeRnd (median=\CorpusRCodeMedian) lines of R code and
\CorpusNativeCodeRnd (median=\CorpusNativeCodeMedian) lines of native code
(C/C++/Fortran). Figure~\ref{fig:corpus} shows these packages, the size of the
dots reflects the project's size in lines of code including both R and native
code, the x-axis indicates the expression code coverage in percents and the
y-axis gives the number of reverse dependencies in log scale. Dotted lines
indicate means. Large (over \CorpusLargePackageTresholdRnd lines of code) and popular (in \CorpusMostDownloadedTreshold most downloaded) packages are annotated.

\begin{figure}[!htb]\centering\includegraphics[width=.9\linewidth]
  {programs-code.pdf}\caption{Programs code size}\label{fig:programs-code}
\end{figure}

\begin{figure}[!htb]\centering\includegraphics[width=.9\linewidth]
  {programs-runtime.pdf}\caption{Programs execution time}\label{fig:programs-runtime}
\end{figure}

These packages come from the Comprehensive R Archive Network
(CRAN\footnote{\url{http://cran.r-project.org}}), the largest repository of R
code with over \CorpusAllCranRnd packages\footnote{CRAN receives about 6 new
  package submissions a day~\cite{Ligges2017}}. Unlike other open code
repositories such as GitHub, CRAN is a curated repository. Each submitted
package must abide to a number of well-formedness rules that are automatically
checked asserting certain quality.

Next to code and tests, R packages also contain examples and vignettes. Examples
are runnable snippets extracted from documentation. Vignettes are tutorial-like
guides into the package functionality. They are written in the literate
programming style using either Latex or Rmarkdown format with embedded code
snippets. Since CRAN is a curated repository, upon submission, it checks all the
runnable code in tests, examples and vignettes and only a successfully running
package is admitted to the archive. We use the same mechanism to extract
runnable code and run it for the dynamic analysis.

In the selected packages, there are \CorpusRunnableCodeRnd lines of runnable
code in examples (\CorpusExamplesCodeRnd), tests (\CorpusTestsCodeRnd) and
vignettes (\CorpusVignettesCodeRnd). Running this code results on average in
\CorpusMeanExprCoverage package code coverage. The code coverage was computed
for each package using

>>>>

\documentclass[conference]{IEEEtran}
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{listings}
\newcommand{\missingNumber}{\textcolor{red}{XX}\xspace}
\newcommand{\missingPercentage}{\textcolor{red}{XX\%}\xspace}
\newcommand{\missingTable}{\textcolor{red}{XXTable}\xspace}
\newcommand{\missingGraph}{\textcolor{red}{XXGraph}\xspace}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{paralist}
\usepackage{xspace}
\usepackage{url}

@@

\newcommand{\eg}{\emph{e.g.},\xspace}
\newcommand{\ie}{\emph{i.e.},\xspace}
\newcommand{\cf}{\emph{cf.}\xspace}
\newcommand{\summary}[1]{({\csname #1Min\endcsname} / {\csname #1Mean\endcsname} / {\csname #1Max\endcsname})}

@@

\begin{figure*}[!tb]\centering\includegraphics[width=\linewidth]
  {corpus.pdf}\caption{Corpus}\label{fig:corpus}
\end{figure*}

@@

We have selected the TOP \CorpusPackages packages based on the number of reverse
dependencies, \ie how many other packages depend on them\footnote{Extracted from
  package metadata using builtin function}. Using this criterion helps us to
choose popular packages which in general should be of a better quality and
should have better code coverage. The resulting set has \summary{CorpusRevdeps}
reverse dependencies (min / mean / max) and consists of \CorpusRCodeRnd lines of
R code and \CorpusNativeCodeRnd lines of native code (C/C++/Fortran).
Figure~\ref{fig:corpus} shows these packages, the size of the dots reflects the
project's size in lines of code including both R and native code, the x-axis
indicates the expression code coverage in percents and the y-axis gives the
number of reverse dependencies in log scale. Dotted lines indicate means. Large
(over \CorpusLargePackageTresholdRnd lines of code) and popular (in
\CorpusMostDownloadedTreshold most downloaded) packages are annotated.

These packages come from the Comprehensive R Archive Network
(CRAN\footnote{\url{http://cran.r-project.org}}), the largest repository of R
code with over \CorpusAllCranRnd packages\footnote{CRAN receives about 6 new
  package submissions a day~\cite{Ligges2017}}. Unlike other open code
repositories such as GitHub, CRAN is a curated repository. Each submitted
package must abide to a number of well-formedness rules that are automatically
checked asserting certain quality. Most relevant for this work is that all of
the \emph{runnable code} is tested and only a successfully running package is
admitted in the archive. This is important as we use this runnable code for the
dynamic analysis of eval calls.

There are three sources of runnable code in a R package. Next to the package
\emph{tests}, there are: \emph{examples} and \emph{vignettes}. Examples are R
code snippets extracted from package documentation into scripts file. Vignettes
are long-form guides into package functionality written using
literate-programming style in \LaTeX\xspace or Rmarkdown with chunks of R code
that can be similarly extracted into R scripts. R provides builtin functions for
tangling the embedded code into files.

The selected packages contain \CorpusPackagePrograms programs with
\CorpusPackageProgramsCodeRnd lines of R code. Most programs are examples
(\CorpusExamplesPrograms) with \CorpusExamplesCodeRnd lines of code, followed by
tests (\CorpusTestsPrograms) with \CorpusTestsCodeRnd lines of code and
vignettes (\CorpusVignettesPrograms) with \CorpusVignettesCodeRnd lines of code.

Running this code results on average in \CorpusMeanExprCoverage package code
coverage. The code coverage was computed for each package using
Covr\footnote{\url{ https://github.com/r-lib/covr}}, the R code coverage tool.
